### 2019-09-11 manuscript.Rmd
manuscript

---
documentclass: jss
author:
  - name: Nan-Hung Hsieh
    affiliation: Texas A&M University
  - name: Brad Reisfeld
    affiliation: Colorado State University 
  - name: Weihsueh A. Chiu
    affiliation: Texas A&M University
    address: >
      4458 TAMU, 
      College Station, TX 77843, USA
    email: \email{wchiu@cvm.tamu.edu}  
title:
  formatted: "\\pkg{pksensi}: R package to apply global sensitivity analysis in physiologically based kinetic models"
  # If you use tex in the formatted title, also supply version without
  plain: "pksensi: R package to apply global sensitivity analysis in physiologically based kinetic models"
  # For running headers, if needed
  short: "**pksensi**: Global SA for PK models in R"
abstract: >
  Sensitivity analysis (SA) is an essential tool for modelers to understand the influence of model parameters on model outputs. It is also increasingly used in developing and assessing pharmacokinetic models. For instance, we recently applied a global SA workflow to reduce the computational burden in the Bayesian Markov Chain Monte Carlo-based calibration process of a physiologically based pharmacokinetic (PBPK) model. Although several SA algorithms and software packages are available, no comprehensive software package exists that allows users to seamlessly solve the PBPK model differential equations, conduct and visualize SA results, and discriminate between the “non-influential” model parameters that can be fixed and those that need calibration. Therefore, we developed an \proglang{R} package, called **pksensi**, to fill this need and make SA more accessible in pharmacological and toxicological research. This package can investigate both parameter uncertainty and sensitivity in pharmacokinetic models, including those with multivariate model outputs, such as multiple time-points and tissue compartments. It also refines the extended Fourier Amplitude Sensitivity Test method for SA by adding a random phase-shift to analyze the statistical variability of the sensitivity index for each model parameter. Furthermore, pksensi includes functions to check the convergence of the SA results. Utilizing pksensi, we successfully reproduced our previously published SA results of human PBPK modeling for acetaminophen and its two primary metabolites. Overall, **pksensi** improves the user experience of performing SA and can create robust and reproducible results for decision making in pharmacokinetic model calibration. 
keywords:
  # at least one keyword must be supplied
  formatted: [Pharmacokinetics, Physiologically based kinetics, Global sensitivity analysis]
  plain:     [Pharmacokinetics, Physiologically based kinetics, Global sensitivity analysis]
preamble: >
  \usepackage{amsmath}
  \usepackage{booktabs}
bibliography: references.bib 
output: rticles::jss_article
# output: word_document
---

@@

Sensitivity analysis (SA) is a mathematical technique to investigate how variations in model parameters affect model outputs [@saltelli2008global]. An increasing number of studies use SA approach to determine which model parameters contribute to high variation in model predictions [@ferretti2016trends]. This technique has also been applied in pharmacology and toxicology research [@loizou2015application; @mcnally2012reconstruction]. Pharmacokinetic models describe the changes in the concentrations or amounts of a substance in the body over time (the various terms for the kinetic models, including "pharmacokinetic", "toxicokinetic", and "biokinetic" models, are used exchangeably here). The goal of SA in pharmacokinetic research is to examine the sensitivity of output variables, such as chemical concentration in blood or tissues, with respect to input parameters, such as anatomical, physiological, and kinetic constants [@mcnally2011workflow]. In addition, this SA approach can guide experimental design and the parameter estimation processes [@zhang2015sobol]. For instance, SA can identify "unidentifiable" parameters that can lead to problems with numerical procedures used in parameter estimation [@chu2010quantitative]. It can be further applied to parameter prioritization and parameter fixing before model calibration [@fphar201800588].

@@

where $x_i$ is the nominal value of the $i$-th parameter, and $\omega_i$ is a vector giving the set of frequencies, one frequency for each parameter, and $\varphi$ is a random phase-shift coefficient ranged from 0 to $2\pi$. The default set of frequencies is according to the study from @saltelli1999quantitative. Through the random phase-shift, the robust result of the sensitivity measurement should be similar across each replication under the same sample size.
For most Sobol'-based SA algorithm, the larger sample size is needed to reduce the probability of having negative estimates. However, FAST estimates are always positive according to its construction.

@@

The R package \pkg{pksensi} provides a useful function to solve ODEs in pharmacokinetic and PK models using analytical and numerical approaches. The solution can perform under the pure \proglang{R} programming environment by linking \pkg{pksensi} with the \pkg{deSolve} package [@JSSv033i09]. However, the most efficient way to achieve high computational speed is through compiled, lower-level languages, such as \proglang{FORTRAN}, \proglang{C}, or \proglang{C++}. The \pkg{pksensi} includes a function to compile and create dynamic-link libraries (.dll) on Windows and shared objects (.so) on Unix-liked systems (e.g., Linux and MacOS). This compiled program can load and execute through the build-in function in \proglang{R}. However, Windows users need to install Rtools to compile the source code by using the GNU GCC compiler. More requirements were detailed in @soetaertr. The \proglang{C} implementation of the example model can be found as an example within \pkg{GNU MCSim} [@JSSv002i09]. The \pkg{pksensi} can also link with \pkg{GNU MCSim} to create the model program, used in solving each system of equations.

@@

Here, we can see (Figure \ref{fig:pk-solvefun-rest}) that \code{vdist} and \code{ke} dominate the plasma concentration before and after 5-hr post chemical intake, respectively. The \code{kgutabs} only plays a crucial role to determine the plasma concentration in the first hour. However, the current result only based on the distribution of model parameters for APAP. Given different input conditions (e.g., range of parameter uncertainty, chemical-dependent parameter value) the result can of course change (result not shown).

@@

Figure \ref{fig:SI-output} shows the relationship between the concentration of the rest of the body (\code{Ccompartment}) and the model parameters at different time points. We can find that \code{kgutabs} and \code{vdist} have higher correlation with \code{Ccompartment} in the beginning (t = 0.01 h) of post-intake duration compared with other parameters, suggesting that the parameters have high impact on the modeling result. The \code{ke} shows a high correlation at the later time period (t = 24.01 h). The \code{BW} does not show any obvious relationship with \code{Ccompartment}.

>>>>

---
documentclass: jss
author:
  - name: Nan-Hung Hsieh
    affiliation: Texas A&M University
  - name: Brad Reisfeld
    affiliation: Colorado State University 
  - name: Weihsueh A. Chiu
    affiliation: Texas A&M University
    address: >
      4458 TAMU, 
      College Station, TX 77843, USA
    email: \email{wchiu@cvm.tamu.edu}  
title:
  formatted: "\\pkg{pksensi}: R package to apply global sensitivity analysis in physiologically based kinetic models"
  # If you use tex in the formatted title, also supply version without
  plain: "pksensi: R package to apply global sensitivity analysis in physiologically based kinetic models"
  # For running headers, if needed
  short: "**pksensi**: Global SA for PK models in R"
abstract: >
  Sensitivity analysis (SA) is an essential tool for modelers to understand the influence of model parameters on model outputs. It is also increasingly used in developing and assessing pharmacokinetic models. For instance, we recently applied a global SA workflow to reduce the computational burden in the Bayesian Markov Chain Monte Carlo-based calibration process of a physiologically based pharmacokinetic (PBPK) model. Although several SA algorithms and software packages are available, no comprehensive software package exists that allows users to seamlessly solve the PBPK model differential equations, conduct and visualize SA results, and discriminate between the “non-influential” model parameters that can be fixed and those that need calibration. Therefore, we developed an \proglang{R} package, called **pksensi**, to fill this need and make SA more accessible in pharmacological and toxicological research. This package can investigate both parameter uncertainty and sensitivity in pharmacokinetic models, including those with multivariate model outputs, such as multiple time-points and tissue compartments. We also refines the extended Fourier Amplitude Sensitivity Test method for SA by adding a random phase-shift to analyze the statistical variability of the sensitivity index for each model parameter. Furthermore, pksensi includes functions to check the convergence of the SA results. Utilizing pksensi, we successfully reproduced our previously published SA results of human PBPK modeling for acetaminophen and its two primary metabolites. Overall, **pksensi** improves the user experience of performing SA and can create robust and reproducible results for decision making in pharmacokinetic model calibration. 
keywords:
  # at least one keyword must be supplied
  formatted: [Pharmacokinetics, Physiologically based kinetics, Global sensitivity analysis]
  plain:     [Pharmacokinetics, Physiologically based kinetics, Global sensitivity analysis]
preamble: >
  \usepackage{amsmath}
  \usepackage{booktabs}
bibliography: references.bib 
output: rticles::jss_article
# output: word_document
---

@@

Sensitivity analysis (SA) is a mathematical technique to investigate how variations in model parameters affect model outputs [@saltelli2008global]. An increasing number of studies use SA to determine which model parameters contribute to high variation in model predictions [@ferretti2016trends]. This technique has also been applied in pharmacology and toxicology research [@loizou2015application; @mcnally2012reconstruction]. Pharmacokinetic models describe the changes in the concentrations or amounts of a substance in the body over time (the various terms for the kinetic models, including "pharmacokinetic", "toxicokinetic", and "biokinetic" models, are used exchangeably here). The goal of SA in pharmacokinetic research is to examine the sensitivity of output variables, such as chemical concentration in blood or tissues, with respect to input parameters, such as anatomical, physiological, and kinetic constants [@mcnally2011workflow]. In addition, this SA approach can guide experimental design and the parameter estimation processes [@zhang2015sobol]. For instance, SA can identify "unidentifiable" parameters that can lead to problems with numerical procedures used in parameter estimation [@chu2010quantitative]. It can be further applied to parameter prioritization and parameter fixing before model calibration [@fphar201800588].

@@

where $x_i$ is the nominal value of the $i$-th parameter, and $\omega_i$ is a vector giving the set of frequencies, one frequency for each parameter, and $\varphi$ is a random phase-shift coefficient ranged from 0 to $2\pi$. The default set of frequencies is according to the study from @saltelli1999quantitative. Through the random phase-shift, the robust result of the sensitivity measurement should be similar across each replication under the same sample size.
For most Sobol'-based SA algorithms, the larger sample size is needed to reduce the probability of having negative estimates. However, FAST estimates are always positive according to its construction.

@@

The R package \pkg{pksensi} provides a useful function to solve ODEs in pharmacokinetic and PK models using analytical and numerical approaches. The solution can perform under the pure \proglang{R} programming environment by linking \pkg{pksensi} with the \pkg{deSolve} package [@JSSv033i09]. However, the most efficient way to achieve high computational speed is through compiled, lower-level languages, such as \proglang{FORTRAN}, \proglang{C}, or \proglang{C++}. The \pkg{pksensi} package includes a function to compile and create dynamic-link libraries (.dll) on Windows and shared objects (.so) on Unix-liked systems (e.g., Linux and MacOS). This compiled program can load and execute through the build-in function in \proglang{R}. However, Windows users need to install Rtools to compile the source code by using the GNU GCC compiler. More requirements were detailed in @soetaertr. The \proglang{C} implementation of the example model can be found as an example within \pkg{GNU MCSim} [@JSSv002i09]. The \pkg{pksensi} can also link with \pkg{GNU MCSim} to create the model program, used in solving each system of equations.

@@

Here, we can see (Figure \ref{fig:pk-solvefun-rest}) that \code{vdist} and \code{ke} dominate the plasma concentration before and after 5-hr post chemical intake, respectively. The parameter \code{kgutabs} only plays a crucial role to determine the plasma concentration in the first hour. However, the current result only based on the distribution of model parameters for APAP. Given different input conditions (e.g., range of parameter uncertainty, chemical-dependent parameter value) the result can of course change (result not shown).

@@

Figure \ref{fig:SI-output} shows the relationship between the concentration of the rest of the body (\code{Ccompartment}) and the model parameters at different time points. We can find that \code{kgutabs} and \code{vdist} have higher correlation with \code{Ccompartment} in the beginning (t = 0.01 h) of post-intake duration compared with other parameters, suggesting that the parameters have high impact on the modeling result. The \code{ke} shows a high correlation at the later time period (t = 24.01 h). The parameter \code{BW} does not show any obvious relationship with \code{Ccompartment}.
